---
title: 'Statistical Inference: Analyzing ToothGrowth Data'
author: "Carl Fredriksson"
output: html_document
---

## Overview

In this part of the project we will analyze the ToothGrowth data from the R datasets package.

## Load the ToothGrowth data

```{r}
library(datasets)
df <- datasets::ToothGrowth
nrow(df)
names(df)
sapply(df, class)
levels(df$supp)
unique(df$dose)
```

The dataset is about the effect of vitamin C on tooth growth in guinea pigs. There are `r nrow(df)` observations of three variables: `r names(df)`.

## Basic summary and data exploration

"len" is a numeric variable that describes tooth length. "len" takes values between `r min(df$len)` and `r max(df$len)`. "supp" is a factor that describes what supplement the guinea pig received. "supp" has two levels: `r levels(df$supp)`. "OJ" means orange juice, "VC" means ascorbic acid. "dose" is a numeric variable that describes the dose of the supplement in milligrams/day. In the dataset "dose" takes the values: `r unique(df$dose)`. Each guinea pig received one of the doses of one of the supplements.

```{r}
summary(df)
```

```{r}
oj <- subset(df, supp=="OJ")
vc <- subset(df, supp=="VC")
par(mfrow=c(1, 2))
hist(oj$len)
hist(vc$len)
```

## Hypothesis testing

I think that "OJ" is a more effective supplement than "VC". To verify this idea I am going to do a hypothesis test where the null hypothesis is that there is no difference between the supplements and the alternative hypothesis is that tooth lengths from "OJ" are longer on average. I am going to use a significance level of $\alpha = 0.05$.

$$
H_0: \mu_{OJ} = \mu_{VC} \\
H_a: \mu_{OJ} > \mu_{VC}
$$
```{r}
t.test(oj$len, vc$len, alternative="greater")$p.value
```

## Conclusion

Since the p-value is smaller than alpha ($p = 0.03031725 < \alpha = 0.05$) I conclude that "OJ" is a more effective supplement for guinea pig tooth growth. I assume that the observations are independent and identically distributed (iid). There are 30 observations for each supplement so that should be enough for the t-test. The data is also not very skewed so that should not be a problem for the t-test.